# (PART\*) Linear mixed models (LMM) in `R` {-}

# Learning objectives

Mixed effects models allow ecologists to overcome a number of limitations associated with traditional linear models.

In this workshop, you will learn when it is important to use a **mixed effects model** to analyze your data. Specifically, we will:

**1.** Describe what are (generalized) mixed effects models

**2.** Identify situations in which the use of mixed effects is appropriate

**3.** Implement basic linear mixed models (LMM) with `R`

**4.** Execute basic generalized linear mixed models (GLMM) with `R`

**5.** Validate, interpret and visualize mixed models with `R`


# Preparing for the workshop

To prepare for this workshop, you must download and install the earliest RStudio and R versions. You must also download the data we will use during this workshop:

-    [R Script](https://qcbsrworkshops.github.io/workshop07/book-en/workshop07-script-en.R)
-    [Fish data](http://qcbs.ca/wiki/_media/qcbs_w7_data.csv)
-    [Arabidopsis data](http://qcbs.ca/wiki/_media/banta_totalfruits.csv)
-    [Inverts data](http://qcbs.ca/wiki/_media/inverts.csv)
-    [glmm_funs](http://qcbs.ca/wiki/_media/glmm_funs.r) (code for use in the GLMM section)


This workshop requires the following `R` packages:

-    [ggplot2](http://cran.r-project.org/web/packages/ggplot2/index.html)
-    [lm4](http://cran.r-project.org/web/packages/lme4/index.html)
-    [AICcmodavg](http://cran.r-project.org/web/packages/AICcmodavg/index.html)
-    [MASS](http://cran.r-project.org/web/packages/MASS/index.html)
-    [vcdExtra](http://cran.r-project.org/web/packages/vcdExtra/index.html)
-    [bbmle](http://cran.r-project.org/web/packages/bbmle/index.html)
-    [DescTools](http://cran.r-project.org/web/packages/DescTools/index.html)

To install them from CRAN, run:

```{r, echo = TRUE, eval = TRUE, message=FALSE, warning=FALSE}
install.packages("ggplot2")
install.packages("lme4")
install.packages("AICcmodavg")
install.packages("MASS")
install.packages("vcdExtra")
install.packages("bbmle")
install.packages("DescTools")
install.packages("remotes")
install.packages("MuMIn")

```

To load these packages, run:

```{r, echo = TRUE, eval = TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(lme4)
library(AICcmodavg)
library(MASS)
library(vcdExtra)
library(bbmle)
library(DescTools)
library(remotes)
library(MuMIn)

```
