
# Final model

The same methods we just saw can be used with a GLMM or LMM to choose between models with various random intercepts and/or random slopes and to choose fixed effects to keep in final model.

Recall the two approaches :

- an **information theoretic approach** (e.g., AICc - Workshop 5)
- a **frequentist approach** (where the significance of each term is evaluated using a likelihood ratio test; LRT, with the `anova()` function)

---

We first code potential models and compare them using AICc:

```{r, echo = TRUE, eval = TRUE}
mpl1 <- glmer(total.fruits ~ nutrient*amd + rack + status +
              (1|X) +
              (1|popu)+
              (1|gen),
data = dat.tf, family = "poisson",
control = glmerControl(optimizer = "bobyqa"))

mpl2 <- update(mpl1, . ~ . - rack) # model without rack
mpl3 <- update(mpl1, . ~ . - status) # model without status
mpl4 <- update(mpl1, . ~ . - amd:nutrient) # without amd:nutrient interaction

aic_tab  <- MuMIn::model.sel(mpl1, mpl2, mpl3, mpl4)
(round(aic_table <- aic_tab[ , c("AICc", "delta", "df")], digits = 2))
```

*Note: We do not cover all possible models above, however, the interaction `amd:nutrient` can only be evaluated if both amd and nutrient (i.e., the main effects) are included in the model.*

Alternatively, we can use `drop1()` and `dfun()` functions to evaluate our fixed effects (`dfun()` converts the AIC values returned by the `drop1()` into $\Delta$AIC values)

```{r, echo = TRUE, eval = TRUE}
dd_LRT <- drop1(mpl1, test = "Chisq")
(dd_AIC <- dfun(drop1(mpl1)))
```

- Strong **rack** effect (dAIC = 55.08 if we remove this variable)
- Effects of **status** and **interaction** term are weak (dAIC < 2)

---

Let's start by **removing the non-significant interaction** term to test main effects of nutrient and clipping

```{r, echo = TRUE, eval = TRUE}
mpl2 <- update(mpl1, . ~ . - amd:nutrient)

mpl3 <- update(mpl2, . ~ . - rack) # pas de rack ou d'interaction
mpl4 <- update(mpl2, . ~ . - status) # pas de status ou d'interaction
mpl5 <- update(mpl2, . ~ . - nutrient) # pas de nutrient ou d'interaction
mpl6 <- update(mpl2, . ~ . - amd) # pas d'herbivorie ou d'interaction
```

Choose the method you want to select the best model:

<ins> Method with AICc </ins>

```{r, echo = TRUE, eval = TRUE}
aic_tab2  <- MuMIn::model.sel(mpl2, mpl3, mpl4, mpl5, mpl6)
(round(aic_table2 <- aic_tab2[ , c("AICc", "delta", "df")], digits = 2))
```

<ins> Method with `drop1()` </ins>

```{r, echo = TRUE, eval = TRUE}
dd_LRT2 <- drop1(mpl2, test = "Chisq")
dd_AIC2 <- dfun(drop1(mpl2))
```

**What are our conclusions ?**

Both the main effects of **nutrient** and **clipping** are strong (large change in AIC of $135.6$ (`mpl5`) and $10.2$ (`mpl6`) if either nutrient or clipping are dropped, respectively).

**Our final model includes :**

Fixed effects
* nutrients
* clipping
* rack

Random effects
* observation-level random effect `(1|X)` 
* populations `(1|popu)` 
* genotypes `(1|gen)`


## Challenge 9 

Use the `inverts` dataset (larval development times (`PLD`) of 74 marine invertebrate and vertebrate species reared at different temperatures and time), answer the following questions:

- 1. What is the effect of feeding type and climate (**fixed effects**) on `PLD`?
- 2. Does this relationship vary among taxa (**random effects**)?
- 3. What is the **best distribution family** for this count data?
- 4. Finally, once you determined the best distribution family, re-evaluate your random and fixed effects.


#### **Challenge 9 Solution:** {-}

```{r, echo = TRUE, eval = TRUE}
# inverts <- read.csv('data/inverts.csv', header = TRUE)
# head(inverts)
# table(inverts$temp, inverts$feeding.type)

# mod.glm <- glm(PLD ~ temp + feeding.type, family = poisson(), data = inverts)
# summary(mod.glm)
# drop1(mod.glm, test = "Chisq")

# boxplot(PLD ~ temp,  data = inverts)
# boxplot(PLD ~ feeding.type,  data = inverts)

# boxplot(predict(mod.glm, type = "response")~inverts$temp)

# plot()

# modglm <- glm(PLD ~ temp + feeding.type, family = poisson(), data = inverts)

# r2 <- ranef(mpl1, condVar = TRUE)
# d2 <- dotplot(r2, par.settings = pp)

# plot(aggregate(PLD ~ taxon, FUN = mean, data = inverts)[,2], aggregate(PLD ~ taxon, FUN = var, data = inverts)[,2], pch = 19)
# abline(a = 0, b = 1, lty = 2)

# mod.glmer <- glmer.nb(PLD ~ temp + feeding.type + (1|taxon), data = inverts)
# mod.glm <- glm.nb(PLD ~ temp + feeding.type, family = poisson(), data = inverts)

# plot(aggregate(PLD ~ taxon, FUN = var, data = inverts)[,2], aggregate(PLD ~ taxon, FUN = mean, data = inverts)[,2])
# abline(a = 0, b = 1, lty = 2 )
```

